---
title: "Multilevel example with blocks"
author: "Bob Verity and Pete Winskill"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Multilevel example with blocks}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo = FALSE}
set.seed(1)
# load the drjacoby package
library(drjacoby)
```

**Purpose:** to compare *drjacoby* infered posterior distribution against the exact analytical solution in a multi-level model implemented via likelihood blocks.

## Model

There are `g` groups with means `mu_1` to `mu_g`. These component means are drawn from a normal distribution with mean 0 and sd 1.0. Within each group there are `n` observations drawn around each component mean from normal disribution with sd 1.0.


```{r}
g <- 5
mu <- rnorm(g)

n <- 5
data_list <- list()
for (i in 1:g) {
  data_list[[i]] <- rnorm(n, mean = mu[i])
}
names(data_list) <- sprintf("group%s", 1:g)
```

Parameters data.frame:

```{r}
L <- 5
df_params <- data.frame()
for(i in 1:5){
  df_params <- df_params |>
    add_parameter(name = paste0("mu_", i), min = -L, max = L, block = c(i, 6))
}
```

## MCMC

```{r cars}
mcmc <- dj$new(data = data_list,
               df_params = df_params,
               loglike = loglike_block,
               logprior = logprior_null,
               chains = 5L)
mcmc$burn(iterations = 1000L, silent = TRUE)
mcmc$sample(iterations = 10000L, silent = TRUE)
```

## Plots

Black = posterior draws

Red = analytical solution to multi-level model

Green = analytical solution assuming independent groups (no second level)

```{r, fig.width=6, fig.height=5}
# extract sampling draws
output_sub <- mcmc$output(phase = "sample")

for (i in 1:5) {
  # get posterior draws
  mu_draws <- output_sub[[sprintf("mu_%s", i)]]
  
  # get analytical solution for this group
  x <- seq(-L, L, l = 1001)
  m <- mean(data_list[[i]])
  fx <- dnorm(x, mean = m * n/(n + 1), sd = sqrt(1/(n + 1)))
  
  # get analytical solution if no multi-level model
  fx2 <- dnorm(x, mean = m, sd = sqrt(1/n))
  
  # overlay plots
  hist(mu_draws, breaks = seq(-L, L, l = 1001), probability = TRUE, col = "black",
       main = sprintf("mu_%s", i))
  lines(x, fx, col = 2, lwd = 4)
  lines(x, fx2, col = 3, lwd = 4)
}
```

## Timing

```{r}
mcmc$timing()
```
