---
title: "Example MCMC Implementation"
author: "Bob Verity and Pete Winskill"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{"Example MCMC Implementation"}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r, echo = FALSE}
# set random seed
set.seed(1)

# load the drjacoby package
library(drjacoby)
```

The likelihood and prior distributions that go into *drjacoby* can be specified by the user either as R functions or as C++ functions. This vignette demonstrates a basic MCMC implementation using both the R and C++ methods, and compares the two in terms of speed.


## Setup

We need the following elements to run *drjacoby*:

1. some data
2. some parameters
3. a likelihood function
4. a prior function

Starting with the data, let's assume that our observations consist of a series of draws from a normal distribution with a given mean (`mu_true`) and standard deviation (`sigma_true`). We can generate some random data to play with:

```{r}
# set random seed
set.seed(1)

# define true parameter values
mu_true <- 3
sigma_true <- 2

# draw example data
x <- rnorm(10, mean = mu_true, sd = sigma_true)
```

For our example MCMC we will assume that we know the correct distribution of the data (i.e. we know the data is normally distributed), and we know that the mean is no smaller than -10 and no larger than 10, but otherwise the parameters of the distribution are unknown. Parameters within *drjacoby* are defined in dataframe format, where we specify minimum, maximum, and initial values of all parameters:

```{r}
# define parameters dataframe
df_params <- data.frame(name = c("mu", "sigma"),
                        min = c(-10, 0),
                        max = c(10, Inf),
                        init = c(5, 1))

print(df_params)
```

In this example we have one parameter (`mu`) that occupies a finite range [-10, 10], and another parameter (`sigma`) that can take any positive value. *drjacoby* deals with different parameter ranges using reparameterisation, which all occurs internally meaning we don't need to worry about these constraints affecting our inference.

Next, we need a likelihood function. This **must** have two input arguments: 1) a vector of parameters, 2) a vector of data, and these **must** be input **in that order**. It also **must** return a single value for the likelihood **in log space**. These constraints on the format of the likelihood function might seem a bit restrictive, but they are needed in order for *drjacoby* to know how to use the function internally. The issue of taking logs is particularly important, as the MCMC will still run even if we forget to take logs, but the results produced will be nonsense!

Inside the likelihood function we can extract individual parameter values from the input vector, and then use these values to calculate the probability of the data. In our example, the likelihood function is quite simple thanks to the `dnorm()` function which can return the density of the normal distribution already in log space:

```{r}
# define log-likelihood function
r_loglike <- function(params, x) {
  
  # extract parameter values
  mu <- params[1]
  sigma <- params[2]
  
  # calculate log-probability of data
  ret <- sum(dnorm(x, mean = mu, sd = sigma, log = TRUE))
  
  # return
  return(ret)
}
```

Finally, we need a prior function. This **must** take a single vector of parameters as input, and it **must** return a single value for the prior probability of those parameters **in log space**. Again, this strict format is required for *drjacoby* to know how to use the prior internally. In our case we will assume a uniform prior on `mu`, and a log-normal prior on `sigma`:

```{r}
# define log-prior function
r_logprior <- function(params) {
  
  # extract parameter values
  mu <- params[1]
  sigma <- params[2]
  
  # calculate log-prior
  ret <- dunif(mu, min = -10, max = 10, log = TRUE) + dlnorm(sigma, meanlog = 0, sdlog = 1.0, log = TRUE)
  
  # return
  return(ret)
}
```

Be careful to ensure that your prior is defined over the same range as specified in the `df_params` dataframe. For example, here our uniform prior for `mu` ranges from -10 to 10, and our log-normal prior for `sigma` ranges from 0 to infinity.


## Running the MCMC

Once we have all the elements above it is straightforward to run a basic MCMC. We simply input the four elements listed above, along with the number of burn-in and sampling iterations By default *drjacoby* prints progress bars to the console to keep you updated on the progress of the MCMC. When running in R markdown we can use the option `pb_markdown = TRUE` to print progress bars in a markdown-friendly way, although you will probably want to leave this option turned off when running interactively (simply delete this argument).

```{r}
# run MCMC
r_mcmc_out <- run_mcmc(data = x,
                       df_params = df_params,
                       loglike = r_loglike,
                       logprior = r_logprior,
                       burnin = 1e3,
                       samples = 1e3,
                       pb_markdown = TRUE)
```

The raw output produced is essentially a large nested list, containing diagnostics, log-likelihoods and parameter values. We can dig down to individual elements of this list using the `$` symbol, which can be combined with auto-complete on most platforms to make life easier. For example, the following line prints the first few `mu` values from from first rung of the the first chain:

```{r}
# get first few mu values
head(r_mcmc_out$chain1$theta_burnin$rung1$mu)
```

Note that the first value is 5.0, which is the initial value for `mu` that we specified in the `df_params` dataframe. The advantages of using [multiple rungs](TODO) and [multiple chains](TODO) are dealt with in later vignettes.


## Exploring outputs and checking MCMC performance

We can produce simple trace plots of output, e.g. the parameter `mu`:

(TODO - should show trace plot over burn-in before showing sampling iterations)

```{r, fig.width=8, fig.height=4}
plot_par(r_mcmc_out, "mu")
```

(TODO - talk about each of these plots)

```{r, fig.width=8, fig.height=4}
plot_par(r_mcmc_out, "sigma")
```



## Using C++ functions

Although *drjacoby* is an R package, under the hood it is running C++ through [Rcpp](https://cran.r-project.org/web/packages/Rcpp/index.html). When we pass an R likelihood function to `run_mcmc()`, as in the example above, the code is forced to jump out of C++ into R to evaluate the likelihood before jumping back into C++. This comes with a computational overhead, which can be avoided by specifying functions directly within C++.

To use C++ functions within *drjacoby* we simply write them directly within R as character strings. As before, there are some constraints on what this function must look like. First, it **must** take as input a `std::vector<double>` of parameters **followed by** a `std::vector<double>` of data. Data **must** be input as doubles, and so data consisting of integer or boolean values should be dealt with as though they are continuous values (for example `TRUE = 1.0`, `FALSE = 0.0`). Second, the function **must** return an object of class `SEXP`. The easiest way to achieve this is to calculate the raw return value as a double, and to use the `Rcpp::wrap()` function when returning to transform to `SEXP`. As before, the value returned should be the likelihood evaluated in **log space**.

One disadvantage of using C++ functions over R functions is that we do not have access to all R's nice distribution functions, for example `dnorm()`, and so functions written in C++ will tend to be mathematically more involved. For example, the `r_loglike()` function defined above can be re-written in C++ as follows:

```{r}
# define cpp log-likelihood function as character string
cpp_loglike <- "SEXP loglike(std::vector<double> params, std::vector<double> x) {
  
  // extract parameters
  double mu = params[0];
  double sigma = params[1];
  
  // sum log-likelihood over all data
  double ret = 0.0;
  for (size_t i = 0; i < sizeof(x); ++i) {
    ret += -0.5*log(2*M_PI*sigma*sigma) - (x[i] - mu)*(x[i] - mu)/(2*sigma*sigma);
  }
  
  // return as SEXP
  return Rcpp::wrap(ret);
}"
```

Similarly, we can define the prior as a C++ function. This function **must** take as input a `std::vector<double>` of parameters, and **must** output a single `SEXP` value representing the prior probability of the parameters in **log space**. The `r_logprior()` function defined above can be re-written in C++ as follows:

```{r}
# define cpp logprior function
cpp_logprior <- "SEXP logprior(std::vector<double> params){
  
  // extract parameters
  double sigma = params[1];
  
  // calculate logprior
  double ret = -log(20.0) - log(sigma) - 0.5*log(2*M_PI*1.0*1.0) - (log(sigma) - 0)*(log(sigma) - 0)/(2*1.0*1.0);
  
  // return as SEXP
  return Rcpp::wrap(ret);
}"
```

With these two functions defined we can run the MCMC exactly the same as before, passing in the new functions:

```{r}
# run MCMC
r_mcmc_out <- run_mcmc(data = x,
                       df_params = df_params,
                       loglike = cpp_loglike,
                       logprior = cpp_logprior,
                       burnin = 1e3,
                       samples = 1e3,
                       pb_markdown = TRUE)
```

You should see that this MCMC runs considerably faster than the previous version that used R functions. There is a short delay initially while the strings are compiled into functions, but this only occurs the first time a function is compiled. We can quantify how much faster the C++ method is using the `microbenchmark` package. In the following we compare the MCMC evaluated over 100,000 iterations using 1) the R likelihood and prior, 2) the C++ likelihood and prior, 3) a mixed method using the C++ likelihood but the R prior.

```{r, echo=FALSE, fig.height = 3, fig.width = 7}
#microbenchmark:::summary.microbenchmark(drjacoby:::bm)
ggplot2::ggplot(drjacoby:::bm, ggplot2::aes(x = expr, y = time / 1e6, fill = expr)) + 
  ggplot2::geom_violin() +
  ggplot2::scale_fill_brewer(palette = 2, type = "qual", guide = FALSE) +
  ggplot2::xlab("") +
  ggplot2::ylab("Time (milliseconds)") +
  ggplot2::coord_flip() +
  ggplot2::theme_bw()
```

```{r, echo = FALSE}
# hidden block, calculate speed-up of C++
median_time <- microbenchmark:::summary.microbenchmark(drjacoby:::bm)$median
speedup <- round(median_time[1]/median_time[3], digits = 1)
```

The pure C++ version was on average around `r speedup` times faster than the pure R version in this example. It is interesting to note that the mixed C++ and R version lies almost exactly between the two pure methods, suggesting that the computational cost of the pure R version is dominated in almost equal measure by calculating the likelihood and the prior. In conclusion, if efficiency is your goal then C++ versions of *both* the likelihood and prior should be used. If ease of programming is more important then R versions of these functions should suffice.

This was a very easy problem, and so required no fancy MCMC tricks. The [next vignette](https://mrc-ide.github.io/drjacoby/articles/metropolis_coupling.html) demonstrates how *drjacoby* can be applied to more challenging problems.
