---
title: "Example MCMC implementation"
author: "Bob Verity and Pete Winskill"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{"Example MCMC implementation"}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r, echo=FALSE}
# set random seed
set.seed(1)

# load the drjacoby package
library(drjacoby)
```

Some intro text...

## Setup

We need the following elements to run **drjacoby**:

1. some data
2. some parameters
3. a likelihood function
4. a prior function

Starting with the data, let's assume that our data are draws from a normal distribution with a given mean (`mu_true`) and standard deviation (`sigma_true`). We will try and estimate these values via MCMC.

```{r}
# define true (unknown) parameters
mu_true <- 3
sigma_true <- 2

# define data
x <- rnorm(100, mean = mu_true, sd = sigma_true)
```

Our parameters must be defined in dataframe format, and we must specify minimum and maximum values, as well as initial values.

```{r}
# define parameters dataframe
df_params <- data.frame(name = c("mu", "sigma"),
                        min = c(-Inf,0),
                        max = c(Inf,Inf),
                        init = c(10,1))

print(df_params)
```

Next we need a likelihood function. This *must* have two input arguments: 1) a vector of parameters, 2) a vector of data. It also *must* return a single value for the likelihood *in log space*.

```{r}
# define loglikelihood function
loglike <- function(theta, x) {
  sum(dnorm(x, mean = theta[1], sd = theta[2], log = TRUE))
}
```

Finall, we need a prior function. This *must* take a single vector of parameters as input, and return a single value for the prior probability of those parameters *in log space*.

```{r}
# define logprior function
logprior <- function(theta) {
  dnorm(theta[1], mean = 0, sd = 10, log = TRUE) +
    dlnorm(theta[2], meanlog = 0, sdlog = 1, log = TRUE)
}
```

## Running the MCMC

Once we have all the elements above, it is straightforward to run the main MCMC. When running the MCMC in R markdown we can use the option `pb_markdown = TRUE` to avoid progress bars printing at every iteration, but when running in ordinary R this should be omitted.

## NOTE - code eval=FALSE past this point, as still fiddling with input types

```{r, eval=FALSE}
# run MCMC
mcmc_out <- run_mcmc(data = x,
                     df_params = df_params,
                     input_type = 1,
                     r_loglike = loglike,
                     r_logprior = logprior,
                     burnin = 1e2,
                     samples = 1e3,
                     chains = 1,
                     pb_markdown = TRUE)
```

We can produce simple trace plots of output

```{r, fig.width=8, fig.height=4, eval=FALSE}
# extract posterior parameter draws
theta <- mcmc_out$chain1$theta_sampling$rung1
mu <- theta$mu
sigma <- theta$sigma

# produce trace plots
plot(mu)
abline(h = mu_true, col = 2)

plot(sigma)
abline(h = sigma_true, col = 2)
```
