---
title: "Installing drjacoby"
author: "Bob Verity and Pete Winskill"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Installing drjacoby}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Installing Rcpp

*drjacoby* relies on the [Rcpp](https://cran.r-project.org/web/packages/Rcpp/index.html) package, which requires the following OS-specific steps:

* Windows
    - Download and install the appropriate version of [Rtools](https://cran.rstudio.com/bin/windows/Rtools/) for your version of R. On installation, ensure you check the box to arrange your system PATH as recommended by Rtools
* Mac OS X
    - Download and install [XCode](http://itunes.apple.com/us/app/xcode/id497799835?mt=12)
    - Within XCode go to Preferences : Downloads and install the Command Line Tools
* Linux (Debian/Ubuntu)
    - Install g++ with
    
        ```{}
        sudo apt-get update
        sudo apt-get install g++
        ```

Irrespective of which system you use above, you should then install and load Rcpp with

```{r, eval = FALSE}
install.packages("Rcpp")
library(Rcpp)
```

You can check the version number to make sure it has properly installed
```{r, eval = FALSE}
packageVersion("Rcpp")
```


## Installing and loading *drjacoby*

Next, in R, ensure that you have the [devtools](https://www.rstudio.com/products/rpackages/devtools/) package installed by running

```{r, eval = FALSE}
install.packages("devtools", repos='http://cran.us.r-project.org')
```

Then install the *drjacoby* package directly from GitHub by running

```{r, eval = FALSE}
devtools::install_github("mrc-ide/drjacoby")
```

If you have any problems installing then please [raise an issue](https://github.com/mrc-ide/drjacoby/issues) on github.

Assuming everything installed correctly, we need to load the package:

```{r}
library(drjacoby)
```

You can test that the package is loaded and working by running the following command, which should produce this output:

```{r}
check_drjacoby_loaded()
```


