---
title: "Installing drjacoby"
author: "Bob Verity and Pete Winskill"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Installing drjacoby}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Installing cpp11

*drjacoby* relies on the [cpp11](https://cran.r-project.org/web/packages/cpp11/index.html) package, which requires [OS-specific setup](https://cran.r-project.org/web/packages/cpp11/vignettes/cpp11.html)

## Installing and loading *drjacoby*

Next, in R, ensure that you have the [devtools](https://www.rstudio.com/products/rpackages/devtools/) package installed by running

```{r, eval = FALSE}
install.packages("devtools", repos='http://cran.us.r-project.org')
```

Then install the *drjacoby* package directly from GitHub by running

```{r, eval = FALSE}
devtools::install_github("mrc-ide/drjacoby")
```

If you have any problems installing then please [raise an issue](https://github.com/mrc-ide/drjacoby/issues) on github.

Assuming everything installed correctly, we need to load the package:

```{r}
library(drjacoby)
```


